% Copyright 2012 Jeffrey Kegler
% This document is licensed under
% a Creative Commons Attribution-NoDerivs 3.0 United States License.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=0.75in,bmargin=0.75in,lmargin=0.75in,rmargin=0.75in}
\usepackage{textcomp}
\usepackage{amstext}
\usepackage{amssymb}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\newcommand{\noun}[1]{\textsc{#1}}
\providecommand*{\perispomeni}{\char126}
\AtBeginDocument{\DeclareRobustCommand{\greektext}{%
  \fontencoding{LGR}\selectfont\def\encodingdefault{LGR}%
  \renewcommand{\~}{\perispomeni}%
}}
\DeclareRobustCommand{\textgreek}[1]{\leavevmode{\greektext #1}}
\DeclareFontEncoding{LGR}{}{}

\newcommand{\lyxmathsym}[1]{\ifmmode\begingroup\def\b@ld{bold}
  \text{\ifx\math@version\b@ld\bfseries\fi#1}\endgroup\else#1\fi}

\makeatother

\usepackage{babel}

\begin{document}

\title{Notes to Aycock and Horspool 2002}

\author{Jeffrey Kegler
\thanks{
Copyright \copyright\ 2012 Jeffrey Kegler.
This document is licensed under
a Creative Commons Attribution-NoDerivs 3.0 United States License.
}
}
\maketitle
\begin{abstract}
The 2002 paper by Aycock and Horspool, on a modification to Jay Earley's
parser, has been essential to my work on general BNF parser. It is
a difficult read for several reasons. First, the paper is dense with
important, but difficult, insights. Second, a reader needs to be familiar
with much previous work, including in the now-neglected field of general
BNF parsing. Finally, there are a number of obscurities and some errors.
This paper is a series of notes and errate intended to address the
last problem.
\end{abstract}

\section{THIS IS A DRAFT}

This paper is in late stages, but not yet finished. It will contain
many errors of the kind that the last drafts should remove.


\section{Introduction}

The 2002 paper by Aycock and Horspool contains a number of obscurities
and some errors. This paper is a series of notes and errata intended
to address the last problem. Notes are not distinguished from errata
in what follows. The difference is a fine one, and it is not clear
that it would serve the reader to know where this author draws the
line in this particular cloud.

The discerning reader will realize that an exercise in nit-picking
like this article is an especially sincere form of praise. Only if
this author truly thought Aycock and Horspool's 2002 paper rewarded
careful and repeated study, would he have collected these notes.


\section{Lemma 5.2 is Incomplete}

The proof of Lemma 5.2 seems incomplete. In case 4 on page 622, the
last sentence begins {}``If $I''$$\in S_{i}'$''. There is no justification
given in the proof for the assertion that $I''$$\in S_{i}'$. A justification
is necessary in order for Lemma 5.2 and Theorem 5.1 to go through.

Case 4 can be shown, and the proof successfully completed, if all
4 cases are wrapped in an induction. This induction could be on completion
depth, where the completion depth of an item is defined as the number
of completion steps ($E_{c}$ or $E'_{c}$) within the current Earley
set needed to add that item to the Earley set. For example, a scanned
item would have zero completion depth.

Lemma 5.3, case 4 relies on Lemma 5.2's case 4 for its justification,
and therefore has the same issue.


\section{Parent Pointers in Theorem 6.1}

In the definitions leading up to Theorem 6.1 on page 624, $l\sqsubset S$
is stated to be equivalent to $[\, A\,\rightarrow\,\lyxmathsym{\textgreek{a}\textbullet\textgreek{b}},\: j]\,\in\, S_{i}$,
where $j$ is never defined. Similarly, $L\sqsubset S_{i}$ is stated
to be equivalent to $l\sqsubset S_{i}$ for all $l\in L$.

LR(0) items never appear alone as member of Earley sets. They are
always in a duple with a parent pointer. Many of the subsequent statements
using the $l\sqsubset S$ and $L\sqsubset S_{i}$ notations are only
true for the correct choice of parent pointer.

The problem can be resolved, and all statements involving the square
subset ($\sqsubset$) notation become true, if it is revised to include
the parent pointer. This can be done by setting $l@j\sqsubset S$
to the equivalent of $[\, A\,\rightarrow\,\lyxmathsym{\textgreek{a}\textbullet\textgreek{b}},\: j]\,\in\, S_{i}$,
and $L@j\sqsubset S_{i}$ to be equivalent to $l@j\sqsubset S_{i}$
for all $l\in L$. The square subset notation is only used in column
one on page 624. If {}``$@j$'' is introduced before the square
subset symbol everywhere it occurs, the expectations for the parent
pointer are made clear. This change in notation leaves Theorem 6.1
true, and its proof correct.


\section{LR(0) State versus LR(0) Item}

On page 624, the notation for LR(0) states and LR(0) items, while
not incorrect, is inconsistent is a way this is very likely to be
confusing. In column one, LR(0) states are always designated with
a capital letter ($L$), while LR(0) items are always designated with
lowercase letters ($l$). But in the second paragraph of column two
on the same page, LR(0) states are designated with the lowercase letter
$l$ , previously reserved for LR(0) items.


\section{Prediction Items are included in LR(0) States}

\noun{Note}: In the second paragraph of page 624, this sentence occurs:
{}``All items $[A\rightarrow\lyxmathsym{\textbullet\textgreek{a}}]$
must be in $l$ too''. As pointed out above, use of $L$ for the
LR(0) state would be more consistent than the use of $l$.

Also, I found it necessary to remind myself why the assertion is true:
By the definition of LR(0) states, whenever an item $[B\rightarrow...\lyxmathsym{\textbullet}A...,k]$
is in an LR(0) state, all prediction items for $A$ must also be in
that same LR(0) state.


\section{\texttt{foreach} over Worklists\texttt{ }}

\noun{Note}: In the pseudocode on page 625 and on page 627 it's important
to realize that, in the \texttt{foreach} loops over the contents of
Earley sets, the Earley sets should be implemented as {}``worklists''.
This is stated on page 625, but is easy to miss. Implementation as
worklists means the the loop must be able to add new items as it proceeds,
and that the iteration includes must include those newly added items.
A naive implementation of a foreach loop would usually not provide
worklist semantics.
\end{document}
