% Copyright 2012 Jeffrey Kegler
% This document is not part of the Marpa or Marpa::R2 source.
% Although it may be included with a Marpa distribution that
% is under an open source license, this document is
% not under that open source license.
% Jeffrey Kegler retains full rights.
\documentclass[12pt,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{listings}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=0.75in,bmargin=0.75in,lmargin=0.75in,rmargin=0.75in}
\usepackage{float}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\newcommand{\noun}[1]{\textsc{#1}}
\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}
\floatname{algorithm}{Algorithm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% Copyright 2012 Jeffrey Kegler
% This document is not part of the Marpa or Marpa::R2 source.
% Although it may be included with a Marpa distribution that
% is under an open source license, this document is
% not under that open source license.
% Jeffrey Kegler retains full rights.

\makeatother

\usepackage{babel}

\begin{document}

\title{Marpa Correctness Proof}


\author{Jeffrey Kegler}
\maketitle
\begin{abstract}
This is a {}``correctness'' proof for the Marpa algorithm. {}``Correctness''
is shown by demonstrating equivalence to the Aycock \& Horspool's
modification of Earley's algorithm in their 2002 paper.

\tableofcontents{}
\end{abstract}

\section{THIS IS A DRAFT}

This paper is in late stages, but not yet finished. It will contain
many errors of the kind that the last drafts should remove.


\section{Introduction}

The paper proves the equivalence of Marpa to Aycock and Horspool's
2002 revision of Earley's algorithm (AH2002). The intent is to establish
confidence that the Marpa algorithm is correct. The reader is assumed
to be familiar with standard grammar notation, LR(0) automata, Earley's
algorithm and Aycock and Horspool's 2002 paper%
\footnote{Aycock and Horspool 2002%
}.


\section{The Two Algorithms}

In this paper, the logic of both the Marpa and the AH2002 algorithm
is divided among five sections
\begin{itemize}
\item \texttt{Initializer}
\item \texttt{Recognizer}
\item \texttt{Earley Set Processing}
\item \texttt{Scanner}
\item \texttt{Completer}
\end{itemize}
The \texttt{Initializer}, \texttt{Recognizer}, \texttt{Scanner} and
\texttt{Completer} are the same for both algorithms. \texttt{Marpa}
and \texttt{AH2002} differ only in their \texttt{Earley Set Processing}.

Pseudocode is provided for \texttt{Recognizer}, \texttt{Scanner},
\texttt{Completer} and each algorithm's \texttt{Earley Set Processing}.
All processing is deterministic. Variables are either global to the
algorithm or local to their section.

Sections {}``call'' other sections. When called, some sections take
arguments, which become input variables in the called section. Input
variables are local to the section, and their semantics is call-by-value.
Despite the similarity to the semantics of subroutines in certain
languages, no implementation is implied. Sections do not return a
value.


\section{Earley Sets}

Each Earley set is an ordered list of Earley items. In AH2002 and
Marpa, every Earley item is a \texttt{(state, parent)} duple, where
\texttt{state} is an AHFA state. AHFA states are states of the split
$\epsilon$-DFA automatons first described by Aycock and Horspool.
In this document a split $\epsilon$-DFA automaton is called an AHFA
(Aycock-Horspool Finite Automaton) and its states are called AHFA
states. Details of AHFA's and their states are not relevant to this
proof and may be found in Aycock and Horspool 2002.

In the \texttt{(state, parent)} Earley item duple, \texttt{parent}
is a parent pointer. A parent pointer is a pointer to an Earley set,
possibly the same one of which the Earley item is a member.


\subsection{Worklists}

In both the AH2002 and Marpa algorithms, the Earley sets need to be
arrays that obey the semantics of worklists. Worklists are arrays
which contain work items. Work items in a work list never disappear,
unless the entire work list goes out of scope. Worklists only change
in one way -- by the addition of new work items at the end of the
work list.

A dynamic array can be implemented so that is never changed except
to grow it at the end. Such an array would provide worklist semantics.


\subsection{AHFA States Versus LR(0) Items}

The most visible difference between AH2002 and the standard Earley
algorithm is in the form of the Earley items. In both AH2002 and the
standard Earley algorithm, Earley items are duples. Also, in both
AH2002 and the standard Earley algorithm, the second element of the
duple is a parent pointer.

But the first element of the duple in AH2002 is not the same as in
the standard Earley algorithm. In the standard Earley algorithm, the
first element of the duple is an LR(0) \textbf{item}. In the AH2002
algorithm the first element of the Earley item is an AHFA \textbf{state}.
An AHFA state is a \textbf{set} of one or more LR(0) items.

Working with sets of LR(0) items instead of single items makes the
AH2002 algorithm more complex, but it also makes it much more efficient
in practice. Earley items in the Marpa algorithm are the same as in
the AH2002 algorithm -- duples of AHFA state and parent pointer.


\section{Initializer}

In both Marpa and AH2002, the Initializer accepts a token stream and
a grammar as input. The token stream is an array of \texttt{n} input
tokens, \texttt{x{[}1{]}, x{[}2{]}, ..., x{[}n{]}}. 

In each algorithm, the Recognizer Main Processing initializes \texttt{S{[}0{]}}
with Earley items of the form \texttt{(start\_state, 0)} where \texttt{start\_state}
is one of the AHFA start states and the parent pointer points to Earley
set 0. There will be one or two start states, depending on the grammar.


\subsection{Pure Functions of the Grammar}

The Initializer creates the functions \texttt{COMPLETED}, \texttt{LHS}
and \texttt{GOTO}. These are pure functions which depend only on the
grammar. These are partial functions with a finite and, in fact, very
limited domain. Since these functions depend only on the grammar,
they may be precomputed. Since they are only defined for a fixed list
of arguments, these functions may be implemented as table lookups.

\texttt{LHS} is a function from a rule to its left hand side symbol.
\texttt{LHS} is defined for every rule of the grammar.

\texttt{COMPLETED} is a function from an AHFA state to a list of the
rules which have completed LR(0) items in that AHFA state. A completed
LR(0) item is one with the dot at the end. The list of rules returned
by \texttt{COMPLETED} may be empty.

\texttt{GOTO} is a transition function from an AHFA state and a symbol
(which may be undefined) to another AHFA state. Most AHFA states only
have transitions on a few of the grammar's symbols, so for many combinations
of \texttt{state} and \texttt{symbol}, the result of \texttt{GOTO(state,
symbol)} will be undefined.


\section{Recognizer}

The Recognizer builds an array of \texttt{n+1} Earley sets, \texttt{S{[}0{]},
S{[}1{]}, ..., S{[}SIZE(x){]}}, where \texttt{SIZE(x)} is the size
of the token stream array, \texttt{x}. In both algorithms, the Earley
sets need to be implemented so they can be treated as worklists.

In both Marpa and AH2002, the Recognizer Main Processing then does
the Earley Set Processing. The Earley Set Processing is performed
for each of the Earley sets from 0 to \texttt{SIZE(x)}, in order.

%
\begin{algorithm}[h]

\begin{lstlisting}
Recognizer Loop: for every (i) in (0 .. SIZE(x)) {
    do Earley Set i Processing
}
\end{lstlisting}


\caption{Recognizer}

\end{algorithm}



\section{The AH2002 Algorithm}

In the proofs, it is convenient to refer to a \texttt{Scanner Loop}
and a \texttt{Completer Loop} for both \texttt{AH2002} and \texttt{Marpa}.
\texttt{Marpa} has separate Scanner and Completer Loops defined in
its Earley Set Processing pseudocode. For \texttt{AH2002}, both the
\texttt{Scanner Loop} and the \texttt{Completer Loop} is defined as
equivalent to the \texttt{Combined Loop}.

\noun{Definition}: \texttt{Scanner Loop @ AH2002 @ Earley Set i Processing}
is defined as \texttt{Combined Loop @ AH2002 @ Earley Set i Processing}.

\noun{Definition}: \texttt{Completer Loop @ AH2002 @ Earley Set i
Processing} is defined as \texttt{Combined Loop @ AH2002 @ Earley
Set i Processing}.

%
\begin{algorithm}[h]

\begin{lstlisting}
Input variable: i

Combined Loop: for every (state, parent) in S[i] {
    do Scanner( state, parent )
    if parent != i {
       do Completer( state, parent )
    }
}
\end{lstlisting}


\caption{AH2002 Earley Set Processing}

\end{algorithm}



\section{The Marpa Algorithm}

The \texttt{Marpa} algorithm is the \texttt{AH2002} algorithm {}``upside-down
and inside-out''. It pulls the scanner and the completer out of the
main loop and gives each its own loop. And \texttt{Marpa }reverses
the order -- the completer is called first to process all items in
an Earley set, including those items added by the completer itself.
After the completer is finished, the scanner is called to process
all items in the Earley set. The scanner never adds items to the current
Earley set, so the current Earley set, \texttt{S{[}i{]}} remains constant
in \texttt{Scanner Loop}.

%
\begin{algorithm}[h]

\begin{lstlisting}
Input variable: i

Completer Loop: for every (state, parent) in S[i] {
    if parent != i {
       do Completer(state, parent)
    }
} # End of Completer Loop

Scanner Loop: for every (state, parent) in S[i] {
    do Scanner(state, parent)
} # End of Scanner Loop
\end{lstlisting}


\caption{Marpa Earley Set Processing}

\end{algorithm}



\section{Scanner}

The \texttt{Scanner} takes three input variables. The first two are
the elements of an Earley item: \texttt{state} is an LR(0) state,
and \texttt{parent} is the index of an Earley set. The third input
variable, i, is the index of the current Earley set.

\texttt{k} and \texttt{nk} are temporaries, local to the subroutine,
which hold AHFA state values returned by \texttt{GOTO}. \texttt{S}
is a global array containing the Earley sets. \texttt{x} is global
array containing the tokens.

%
\begin{algorithm}[h]

\begin{lstlisting}
Input variables: state, parent, i

set k = GOTO(state, x[i+1])
if state k is not defined {
    return
}
add (k, parent) to S[i+1] at end, if it is not already present
set nk = GOTO(k, undefined)                                   
if state nk is not defined {
    return
}
add (nk, i+1) to S[i+1] at end, if it is not already present
return
\end{lstlisting}


\caption{Scanner}

\end{algorithm}



\section{Completer}

The \texttt{Completer} takes three input variables. The first two
are the elements of an Earley item: \texttt{state} is an LR(0) state,
and \texttt{parent} is the index of an Earley set. The third input
variable, i, is the index of the current Earley set.

\texttt{k} and \texttt{nk} are temporaries, local to the subroutine,
which hold LR(0) state values returned by \texttt{GOTO}. \texttt{rule}
is a rule, returned by \texttt{COMPLETED}. \texttt{S} is a global
array containing the Earley sets.

%
\begin{algorithm}[h]

\begin{lstlisting}
Input variables: state, parent, i

foreach rule in COMPLETED(state) {
    foreach ( pstate, pparent ) in S[parent] {
    set k = GOTO( pstate, LHS(rule) )
    if k is defined {
         add ( k, pparent ) to S[i] at end, if it is not already present
         set nk = GOTO( k, undefined )
         if nk is defined {
              add ( nk, i ) to S[i] at end, if it is not already present
         }
     } # if k is defined
}  # foreach rule
\end{lstlisting}


\caption{Completer}

\end{algorithm}



\section{The Proof: Short Version}

After, Initialization, the only changes to globals are to the Earley
sets, and these occur in the Earley Set Processing. The Processing
of Earley set i can change S{[}i{]} (called the current Earley set)
and S{[}i+1{]} (the next Earley set). The proof is organized around
these facts.

First, two lemmas are proved showing that if, before each Earley Set
Processing step, the Earley sets in both algorithms are identical,
then, after the Earley Set Processing Step, the current Earley sets
and the next Earley sets remain identical .


\subsection{The Current Earley Set Lemma}

\noun{Statement of the Lemma}: Assume that, before each Earley set
is processed, the Earley sets in AH2002 are identical to the Earley
sets in Marpa. Then, after Earley Set i Processing, S{[}i{]} in AH2002
is identical to S{[}i{]} in Marpa.

\noun{Proof}: In the Earley Set Processing, S{[}i{]} will only be
changed by the Completer. The Completer can be treated as a pure deterministic
functions whose inputs are its input variables (state, parent and
i), the current Earley set itself (S{[}i{]}) and the pure functions
of the grammar (GOTO, LHS and COMPLETED). GOTO, LHS, COMPLETED do
not change after Initialization and the input variable i will not
change throughout the Earley Set Processing, so they may treated as
constants for this purpose.

The input variables (state, parent) and the current Earley set (S{[}i{]})
do change but, in both AH2002 and Marpa, the calls to the Completer
will be with same values for these and in the same sequence. This
can be shown by induction on the calls to the Completer.

There may be zero such calls, and if this is the case, S{[}i{]} after
Earley Set i Processing will clearly be unchanged from S{[}i{]} before
Earley Set i Processing. Assume there have been n calls to the completer.
By the induction hypothesis, after these n calls, S{[}i{]} is identical
in both algorithms. The input variables state and parent depend only
on S{[}i{]}, and on i. i was already shown to be constant during the
Earley Set i Processing. S{[}i{]} is identical in both algorithms
by the induction hypothesis. The Completer is the same in both algorithms.
Therefore, since the Completer and its inputs are identical, and the
Completer is deterministic, the results of the Completer will be the
same in both algorithms.

The induction above assumed that there is a n+1'th call to the Completer
in both algorithms. This actually depends on the domain for the Completer
Loop. But this domain is also the current Earley set, S{[}i{]}. So
by the same induction as above, it is not possible there will be a
n+1'th call to the Completer in one algorithm, without there being
an n+1'th call to the Completer in the other algorithm. In other words,
either there will be a n+1'st call to the Completer in both algorithms,
or the n'th call to the Completer in both algorithms will have been
the final one.

This completes the induction and shows that, after the Earley Set
i Processing, the current Earley sets (S{[}i{]}) in both algorithms
must be identical to each other. QED.


\subsection{The Next Earley Set Lemma}

\noun{Statement of the Lemma}: Assume that, before each Earley set
is processed, the Earley sets in AH2002 are identical to the Earley
sets in Marpa. Then, after Earley Set i Processing, S{[}i+1{]} in
AH2002 is identical to S{[}i+1{]} in Marpa.

\noun{Proof}: First, recall the Current Earley Set Lemma. By it, we
know that \textbf{after} Earley Set i Processing, the \textbf{current}
Earley set (S{[}i{]}) is identical in both AH2002 and Marpa. We will
make use of this fact below.

In the Earley Set Processing, S{[}i+1{]} will only be changed by the
Scanner. The Scanner can be treated as a pure deterministic function
whose inputs are its input variables (state, parent and i), the next
Earley set itself (S{[}i+1{]}), the token stream (x) and a pure function
of the grammar (GOTO). GOTO and x do not change after Initialization
and the input variable i will not change throughout the Earley Set
Processing, so they may treated as constants for this purpose.

The input variables (state, parent) and the next Earley set (S{[}i+1{]})
do change but, in both AH2002 and Marpa, the calls to the Scanner
will be with same values for these, will be made in the same sequence,
and will have the same results. This can be shown by induction on
the calls to the Scanner.

There may be zero such calls, and if this is the case, S{[}i+1{]}
after Earley Set i Processing will clearly be unchanged from S{[}i+1{]}
before Earley Set i Processing. Assume there have been n calls to
the Scanner. By the induction hypothesis, after these n calls, S{[}i+1{]}
is identical in both algorithms.

The input variables state and parent are the n+1'st work item in the
current Earley set S{[}i{]}. Call this work item S{[}i{]}{[}n+1{]}.
As a work item, once S{[}i{]}{[}n+1{]} is added to a work list it
is never deleted and never changes. Since by the Current Earley Set
Lemma, S{[}i{]} in AH2002 will be identical to S{[}i{]} in Marpa after
Earley Set i Processing, if S{[}i{]}{[}n+1{]} exists in both algorithms
it must be identical.

S{[}i+1{]} is identical in both algorithms by the induction hypothesis.
The Scanner is the same in both algorithms. Therefore, since the Scanner
and its inputs are identical, and the Scanner is deterministic, the
results of the Scanner will be the same in both algorithms.

The induction above assumed that there is a n+1'th call to the Scanner
in both algorithms. This actually depends on the domain for the Scanner
Loop. This domain is the current Earley set, S{[}i{]}.

But what if there was an S{[}i{]}{[}n+1{]} in AH2002 but no S{[}i{]}{[}n+1{]}
in Marpa, or vice versa? We can show this won't happen, by a reduction
to absurdity. Suppose, just after the n'th call to the Scanner, S{[}i{]}{[}n+1{]}
exists in AH2002, but not in Marpa. By the Current Earley Set Lemma
S{[}i{]} in both algorithms is identical after the Earley Set i Processing.
But if S{[}i{]}{[}n+1{]} does not exist for the Scanner Loop in Marpa,
examination of the pseudocode will show that S{[}i{]}{[}n+1{]} will
not exist in Marpa at the End of Earley Set i Processing either. On
the other hand, if S{[}i{]}{[}n+1{]} exists at the End of Earley Set
i Processing in AH2002, it will still exist at the End of Earley Set
i Processing in AH2002. This means that S{[}i{]} at the End of Earley
Set i Processing in AH2002 will not be the same as S{[}i{]} at the
End of Earley Set i Processing in Marpa.

But, by the Current Earley Set Lemma, at the End of Earley Set i Processing,
S{[}i{]} in each of the two algorithms must be identical to S{[}i{]}
in the other. This means that the assumption for the reduction to
absurdity in the paragraph above must be false. For the reduction
to absurdity, we assumed that, just after the n'th call to the Scanner,
there was an S{[}i{]}{[}n+1{]} in AH2002, but not in Marpa. Therefore
if, just after the n'th call to the Scanner, there is an S{[}i{]}{[}n+1{]}
in the AH2002, there must also be one in Marpa.

The same reduction to absurdity works in reverse, and shows the if,
just after the n'th call to the Scanner, there is an S{[}i{]}{[}n+1{]}
in Marpa, there must also be one in AH2002. Putting the two together,
just after the n'th call to the Scanner, there is an S{[}i{]}{[}n+1{]}
in AH2002 if and only if there is an S{[}i{]}{[}n+1{]} in Marpa.

This means that it is not possible there will be a n+1'th call to
the Scanner in one algorithm, without there being an n+1'st call to
the Scanner in the other algorithm. In other words, either there will
be a n+1'st call to the Scanner in both algorithms, or the n'th call
to the Scanner in both algorithms will have been the final one.

This completes the induction and shows that, after the Earley Set
i Processing, the next Earley sets (S{[}i+1{]}) in both algorithms
must be identical to each other. QED.


\subsection{Showing Equivalence}

We say that AH2002 and Marpa are equivalent if, when they are run
on the same token streams and grammars, they produce the same Earley
sets. Given the Current Earley Set Lemma and the Next Earley Set Lemma,
the proof follows quickly. The proof is by induction on the Earley
Set Processing.

Initialization is the same in AH2002 and Marpa, and the input to the
algorithms (their tokens streams and grammars) are assumed to be identical.
Initialization is deterministic, so before the Processing of Earley
Set 0, all globals are identical.

Suppose before the Earley Set n Processing, all globals are identical.
We know from examination of the pseudocode that, during Earley Set
n Processing, the only globals which change are S{[}n{]} and S{[}n+1{]}.

We have from the induction hypothesis that, before Earley Set n Processing,
S{[}n{]} in AH2002 is identical is S{[}n{]} in Marpa. From this and
the Current Earley Set Lemma, we can conclude that, after Earley Set
n Processing, S{[}n{]} in each of the algorithms will be identical
to S{[}n{]} in the other algorithm.

We have from the induction hypothesis that, before Earley Set n Processing,
S{[}n+1{]} in AH2002 is identical is S{[}n+1{]} in Marpa. From this,
the Current Earley Set Lemma and the Next Earley Set Lemma, we can
conclude that, after Earley Set n Processing, S{[}n+1{]} in each of
the algorithms will be identical to S{[}n+1{]} in the other algorithm.

Above we have that:
\begin{itemize}
\item All globals remains constant during Earley Set Processing except S{[}n{]}
and S{[}n+1{]}.
\item After Earley Set n Processing, S{[}n{]} in each of the algorithms
will be identical to S{[}n{]} in the other algorithm.
\item After Earley Set n Processing, S{[}n+1{]} in each of the algorithms
will be identical to S{[}n+1{]} in the other algorithm.
\end{itemize}
Combining these three, we have that if, after Earley Set n Processing,
all globals in AH2002 will be identical to the corresponding global
in Marpa. The point after Earley Set n Processing is the same as the
point before Earley Set n+1 Processing. This proves the induction.

The Recognizer will process l+1 Earley sets, where l is the length
of the token stream. From the induction, we know that, after the processing
of the l+1 Earley sets, the Earley sets in Marpa will be identical
to those in AH2002. This is the equivalence we were after.

QED.


\section{Why a Detailed Proof?}

The rest of this paper repeats the  proof already given, more carefully
and in more detail. It describes some new notations, useful for detailed
proofs of this kind. Readers, if satisfied with what they have seen
so far, will not necessarily want to read any further.

The form of the short proof above is similar to that of the correctness
proofs as they usually appear in journals. Proofs in this form are
problematic. English-language descriptions of program locations, even
when carefully written, are verbose. Even so, these descriptions are
always in danger of being ambiguous.

More importantly, the reader can not easily confirm that all crucial
details are being established. For example, in this proof, the existence
or non-existence of loop passes and worklist items at certain points
is as crucial as any issue in the proof, but very easy to overlook.

Short, open-form, English-language proofs of this kind, while they
dominate the academic literature, are error-prone and hard to read.
In the journals, perhaps reasons of space force their acceptance.
Page-count in academic journals is a scarce resource, inelastic to
the demands of academics. Detailed proofs would perhaps double the
size of the corresponding articles. Doubling the page count means
halving the number of articles published. Everyone in the field realizes
that that one or more of the articles left unpublished might well
be his own.

Web space is not nearly so expensive as space in printed academic
journals. The rest of this paper contains the detailed working out
of the correctness proof just given in short form. The short proof
was written after, and based on, the detailed proof.


\section{Notation for the Detailed Proof}


\subsection{Grafs}

Most detailed proofs are in table form, with one equation per line.
Equations are right aligned and their explanations left aligned. Program
correctness proofs require a degree of explanation which easily overwhelms
the one-line-per-equation format.

This proof is organized with one equation per paragraph, usually abbreviated
{}``graf''. The first sentence is an equation, and the others justify
its derivation or provide explanations.


\subsection{Time-Sets}

This proof I introduce time-sets notation. Time-set notation can be
looked at as an abbreviation of English, or as a semi-formalized,
notation. I hope it is an improvement in precision. I would also claim
it is more readable than its expansion into ordinary English would
be.

A key to this notation is that the idea of sequence is an afterthought
in time-set notation. Time-sets are unordered sets of point in the
execution -- time-points. Statements about their sequence -- which
time-point is before what other time-point, can obviously be important
in analyzing programs. But I think premature introduction of the sequencing
into the semantics is counterproductive.

The notation for locations in the code uses sets of time-points. The
points in time, or time-points, are points in the processing -- not
locations in the pseudocode. Every operation of any kind separates
time-points, so many time-points can correspond to the same pseudo-code
location. Additionally, if a lexical location in the pseudocode is
passed through several times, as it would be in a loop, each pass
creates a new set of time-points for that location.

For time-points to be separate, some operation must happen between
them. This means that sometimes the same time-point can be identified
with two different lexical locations in the pseudocode, and have two
different names. This occurs frequently in loops. More on this below.


\subsection{Names of Sets}


\subsubsection{Algorithms}

Each algorithm's name is used for the time-set of all the time-points
in that algorithm. So AH2002 is the name of the time-set of all the
time-points in the execution of the AH2002 algorithm. Marpa is the
name of the time-set of all the time-points in the execution of the
Marpa algorithm. Start of Marpa is the time point before any operations
in the Marpa algorithm. End of Marpa is the time-point after any operations.


\subsubsection{Pseudocode Sections}

Each pseudocode section's name is used for the time-set of all time-points
in that pseudo-code section, including all time-points in either algorithm.
Recognizer is the name of the time-set of all time-points in the Recognizer,
whether for AH2002 or Marpa. When it is useful to distinguish the
@-notation can be used: \texttt{Recognizer @ AH2002} or \texttt{Recognizer
@ Marpa}.

\texttt{Start of Recognizer} is the set of the time-points before
any operation in the Recognizers, including both algorithms. \texttt{End
of Recognizer @ Marpa} is the singleton set containing the time-point
after all operations in the Marpa algorithm.

Pseudocode sections can be called multiple times. For example, \texttt{Scanner
@ Marpa} refers to all time-points in any call of \texttt{Scanner}
in the \texttt{Marpa} algorithm.

Where it is useful to specifically distinguish one call to a section,
this can be done in the time-set's name. For example, the call the
\texttt{Earley Set Processing} in the \texttt{AH2002} algorithm for
Earley Set \texttt{s} is \texttt{Earley Set s Processing @ AH2002}.


\subsection{Loops}

Loops have not only a \texttt{Start} and \texttt{End}, but a \texttt{Top}
and a \texttt{Bottom}. The \texttt{Start of Loop L} is before execution
of any instructions in the loop. The \texttt{End of Loop L} is after
execution of any instructions. The \texttt{Top of Loop L(c)} is before
the execution of pass \texttt{c} through \texttt{Loop L}. The \texttt{Bottom
of Loop L(c)} is after the execution of pass \texttt{c }through Loop
\texttt{L}.

Passes through loops are numbered starting at 0. The final test which
terminates a loop is considered a pass, but a pass which does not
execute any instructions inside the loop. A loop over the numbers
from 0 to 10 will have 12 passes, with pass 0 being the first pass,
and pass 11 being the last pass. Pass 0 will have the loop variable
with the value 0. In pass n, the value of the loop variable will be
n, unless n is the last pass. In the last pass, only the test is performed
-- the fails, none of the instructions inside the loop are executed,
and the loop terminates. Where z is the last pass, there is still
considered to be a both \texttt{Top of Loop(z)} and \texttt{Bottom
of Loop(z)}.

\noun{Definition}: Let L be the name of a loop. \texttt{DOMAIN(L)}
is the array over which loop \texttt{L} iterates.

In this definition, it is important that L is not a timeset -- it
is the name of a loop. Also, it is important to note the value of
\texttt{DOMAIN} can change if the loop is over a worklist.

\noun{Definition}: Let L be the name of a loop. Then,\texttt{ PASS(p,
L)} $\Leftrightarrow$ \texttt{0 <= p <= SIZE(DOMAIN(L) @ End of L))}.

Informally, \texttt{PASS(p,L)} is true if \texttt{p} is the number
of a valid pass of \texttt{L}. If \texttt{D == DOMAIN(L)}, then the
loop variable for pass \texttt{p} is \texttt{D{[}p{]}}.

For any loop \texttt{L, PASS(p, L)} $\wedge$ \texttt{PASS(p+1, L)}
$\Leftrightarrow$ \texttt{(Bottom(p) of L == Top(p+1) of L}).

\noun{Definition}:\texttt{ FINAL(L) == SIZE( DOMAIN(L) @ End of L
)}


\subsection{Operations on Time-Sets}

Subset: \texttt{S1 $\subseteq$ S2}

The notation remains semi-formal. @-notation is often written so that
it's intended to be distributed over other notation. For example:

\texttt{(SIZE(x @ A) == SIZE(x @ B)) @ P}

is an abbreviation for

\texttt{SIZE(x @ A @ P) == SIZE(x @ B @ P)}

The general idea is that in \texttt{(expression) @ P}, where \texttt{expression}
is not a term or another time-set, \texttt{P }should be {}``pulled
inside'' \texttt{expression}. Rules for this are not explicitly formalized,
but use of this syntax is limited to simple cases where it should
be clear what is intended -- certainly far more clear than the usual
English language description would be.


\section{Work List Lemma}

Statement of the Lemma:

Let \texttt{L} be a work list iteration.

Let \texttt{W} be a work list.

Then \texttt{0 <= j <= SIZE( W @ End of L )$\Rightarrow$PASS( j,
L )}

And \texttt{FINAL( SIZE( W @ End of L ), L)}

\noun{Proof}: The proof is from the definitions of loop and work list.
\texttt{L} is defined as being an iteration over every item in \texttt{W},
even those added during \texttt{L}, so that any item added to \texttt{W}
before \texttt{End of L} must be included in the iteration.

Only one operation is allowed on \texttt{W} -- adding items to the
end. This operation does not move or delete any items, so no item
in \texttt{W} will ever be moved or deleted. That means any iterated
item will be present in \texttt{W @( End of L)}, and in the same position
as when it was iterated. Therefore, if there is a \texttt{W{[}j{]}@(End
of L)}, the loop variable for pass j will be \texttt{W{[}j{]}}.

\noun{QED.}


\section{Globals}


\subsection{The Globals Assumption}

For the purpose of the proof in this paper, it is assumed that AH2002
and Marpa are initialized with the same input token stream and grammar.


\subsection{The Globals Lemmas}

Statement of Lemma 1: Let G be a global.

Then G @ Start of Recognizer @ AH2002 == G @ Start of Recognizer @
Marpa.

Proof: From the definitions, we know that the Initialization code
is shared by the two algorithms. From the Globals Assumption, we have
that the results of Initialization depends only on the input tokens
and the grammar.

QED.

\noun{Statement of Lemma} 2: Let G be global not in the Earley sets.

Let P be a time-set, P $\subseteq$Recognizer.

Let Q be a time-set, P $\subseteq$Recognizer.

Then For all G, where G is a global not in the Earley sets, G @ P
== G @ Q.

Proof: The pseudocode and the definition of the Recognizer show that
the only globals that change in the Recognizer are those in the Earley
Sets.

\noun{Statement of Lemma 3}: The Earley sets are changed only in the
the following ways:
\begin{itemize}
\item S{[}0{]} is changed in Initialization;
\item S{[}i{]} is changed in Completer @ Earley Set i Processing; and
\item S{[}i+1{]} is changed in Scanner @ Earley Set i Processing.
\end{itemize}
\noun{Proof}: From the pseudocode and the definition of Initialization.

QED


\section{The Detailed Proof}


\subsection{AH2002 Halting Axiom}

In this paper, AH2002 will be assumed to halt for all finite grammars
and input token streams. AH2002 in their paper do not explicitly prove
that this is true in general, but it is at the very least a safe conjecture.


\subsection{The Completer Lemma}

In this lemma A may be either Marpa or AH2002.

\noun{Statement of the Lemma}:

Assume p is a pass in the Completer Loop.

Assume that \texttt{((state, parent, i) @ AH2002 == (state, parent,
i) @ Marpa) @ Top of Completer Loop(p)}.

Assume \texttt{(S{[}i{]} @ A == S{[}i{]} @ B) @ Top of Completer Loop(p)}.

Then \texttt{(S{[}i{]} @ A == S{[}i{]} @ B) @ Bottom of Completer
Loop(p)}.

\noun{Starting the proof}:

\texttt{0 <= p <= SIZE(S{[}i{]} @ End of Completer Loop)}. By the
Loop Range Property.

The proof proceeds in two cases -- the Final Case \texttt{(p == SIZE(S{[}i{]}
@ End of Completer Loop))} and the Kernel Case \texttt{(0 <= p < SIZE(S{[}i{]}
@ End of Completer Loop))}.

\noun{Kernel case}:

\texttt{((state, parent, i) @ AH2002 == (state, parent, i) @ Marpa)
@ Top of Completer Loop(p)}. By assumption in the statement of the
lemma.

\texttt{((state, parent, i) @ AH2002 == (state, parent, i) @ Marpa)
@ Top of Completer @ Completer Loop(p)}. From the statement just previous
and inspection of the psuedocode. These variables are never changed
between \texttt{Top of Completer Loop(p)} and \texttt{Top of Completer
@ Completer Loop(p)}.

For \texttt{G} in LHS, GOTO, COMPLETED, \texttt{(G @ Marpa == G @
AH2002) @ Top of Completer @ Completer Loop(p)}. By the Globals Lemma.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Start of Completer
@ Completer Loop(p}). By assumption in the statement of the lemma.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ End of Completer
@ Completer Loop(p)}. The Completer pseudocode shows that S{[}i{]}
@ End of Completer is a deterministic function of S{[}i{]} itself,
the Completer's input variables and these globals: LHS, GOTO, COMPLETED.
By the previous 3 grafs, these are the same in both algorithms. The
Completer is also the same in both algorithms. Therefore the effects
of the Completer must be the same in both algorithms.

\texttt{(S{[}i{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom of Completer
Loop(p)}. From the pseudocode. The current Earley set is never changed
between \texttt{Bottom of Completer @ Completer Loop(p)} and \texttt{Bottom
of Completer Loop(p)}.

\noun{This proves the Kernel case}.

\noun{Final case}:

\texttt{FINAL(p, Completer Loop)}.

\texttt{(S{[}i{]} @ Marpa == S{[}i{]} @ AH2002) @ Bottom of Completer
Loop(p)}. From the previous graf and an assumption in the statement
of this lemma. By the definition of the final pass, it is a no-op.
The Completer is not called and \texttt{S{[}i{]}} remains to its value
at the \texttt{Top Completer Loop(p)}.

\noun{This proves the final case}.

With the two cases proved, the Lemma is shown.

QED.


\subsection{The Completer Loop Lemma}

When a task property Lemma is used in this Lemma as a justification,
and no task is mentioned, the task is the Completer pseudocode, the
loop is the Completer Loop and the offset is 1, and the Completer
Task Lemma is to be considered part of the justification.

\noun{Statement of the Lemma}:

Let \texttt{p} be a pass in \texttt{Completer Loop @ Earley Set i
Processing}.

Assume for all globals G not in the Earley sets, \texttt{(G @ Marpa
== G @ AH2002) @ Earley Set i Processing)}.

Assume \texttt{(i @ Marpa == i @ AH2002) @ Earley Set Processing}.

Assume \texttt{(S{[}i{]} @ Marpa == S{[}i{]} @ AH2002) @ Top(p) of
Completer Loop}(p).

Then \texttt{(S{[}i{]} @ Marpa == S{[}i{]} @ AH2002) @ Bottom of Completer
Loop(p)}.

\noun{Proof:}

\texttt{(S{[}i{]}{[}p{]} @ AH2002 == S{[}i{]}{[}p{]} @ Marpa) @ Top
of Completer Loop(p)}. By assumptions in for the lemma.

\texttt{((state, parent) @ AH2002 == (state, parent) @ Marpa) @ Top
of Completer Loop(p)}. From the statement just previous. \texttt{(state,
parent) == S{[}i{]}{[}p{]}} because both are the loop variable.

\texttt{(i @ AH2002 == i @ Marpa) @ Top of Completer Loop (p)}. From
an assumption for this lemma.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Top of Completer
Loop(p)}. An assumption for this lemma.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Bottom of Completer
Loop(p)}. By the Completer Lemma with \texttt{p} set to its value
in this lemma, and by the three previous grafs.

\noun{This proves the lemma.}


\subsection{The Current Earley Set Lemma}

\noun{Statement of the Lemma}:

Assume \texttt{0 <= i <= SIZE(x)}.

Assume \texttt{(S @ AH2002 == S @ Marpa) @ Start of Earley Set i Processing}.

Then, \texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ End of Earley
Set i Processing}.

\noun{Starting the Proof}:

\noun{Sublemma}:

Assume \texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Top of Completer
Loop(0)}.

Assume \texttt{p <= SIZE(S{[}i{]}) @ AH2002 @ Bottom of Completer
Loop(p)} or \texttt{p <= SIZE(S{[}i{]}) @ Marpa @ Bottom of Completer
Loop(p)}.

Then \texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Bottom of Completer
Loop(p)}.

\noun{Note}: In this sublemma, it is assumed in the statement of the
sublemma that pass p exists for one algorithm, but not for the other.
Loop operations are not defined for non-existent passes, and it must
be proved that pass p exists for both AH2002 and Marpa.

\noun{Starting the proof}:

\noun{Induction Property}: \texttt{(S{[}i{]} @ A == S{[}i{]} @ Marpa)
@ Bottom of Completer Loop(p)}, where the induction is on \texttt{p}.

\noun{Induction} \noun{Basis}: \texttt{(S{[}i{]} @ A == S{[}i{]} @
Marpa) @ Bottom of Completer Loop(0)}. By the assumptions in the statement
of the sublemma and the Completer Loop Lemma with \texttt{p := 0}.
Pass 0 exists for every loop, by definition.

\noun{This proves the induction basis.}

\noun{Induction hypothesis}:

Assume \texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Bottom of
Completer Loop(n)}.

\noun{Inductive step}: 

Note: Next it must be shown that \texttt{n+1} is a pass in both AH2002
and Marpa. 

\noun{Subsublemma}: Let \texttt{(A,B)} be either \texttt{(AH2002,
Marpa)} or \texttt{(Marpa, AH2002)}.

Assume \texttt{PASS(n+1, Completer Loop @ A)}.

Then \texttt{PASS(n+1, Completer Loop @ B)}.

Starting the Proof:

\texttt{0 <= n < SIZE(DOMAIN(Completer Loop) @ Bottom of Completer
Loop(n) @ A))}. By the assumption for the sublemma, and the definition
of PASS.

\texttt{DOMAIN(Completer Loop)} is \texttt{S{[}i{]}.} By the definition
of \texttt{DOMAIN}.

\texttt{0 <= n+1 <= SIZE(S{[}i{]} @ Bottom of Completer Loop(n) @
A)}. By the previous 2 grafs. {[} \noun{Current Set Size Graf}. {]}

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Bottom of Completer
Loop(n).} By the Induction Hypothesis. The assumption for the case
is that f == 0.

\texttt{(S{[}i{]} @ A == S{[}i{]} @ B) @ Bottom of Completer Loop(n).}
By the previous, substituting A and B, and transposing the equality
if necessary.

\texttt{0 < n+1 <= SIZE(S{[}i{]} @ Bottom of Completer Loop(n) @ B)}.
By the previous graf and Current Set Size Graf.

\texttt{0 < n+1 <= SIZE(DOMAIN(Completer Loop) @ Bottom of Completer
Loop(n) @ A))}. By the previous graf and the definition of \texttt{DOMAIN}.
The domain of the Completer Loop is \texttt{S{[}i{]}}.

\texttt{PASS(n+1, Completer Loop @ B)}. By the previous graf and the
definition of PASS.

\noun{This proves the subsublemma}.

\texttt{PASS(n+1, Completer Loop @ AH2002) $\Rightarrow$ PASS(n+1,
Completer Loop @ Marpa)}. By the sublemma, with \texttt{(A, B) :=
(AH2002, Marpa)}.

\texttt{PASS(n+1, Completer Loop @ Marpa) $\Rightarrow$ PASS(n+1,
Completer Loop @ AH2002)}. By the sublemma, with \texttt{(A, B) :=
(Marpa, AH2002)}.

Pass Equivalence Graf: \texttt{PASS(n+1, Completer Loop @ AH2002)
$\Leftrightarrow$PASS(n+1, Completer Loop @ Marpa)}. By the previous
two grafs.

\texttt{PASS(n+1, Completer Loop @ AH2002)}. By the assumption for
Sublemma 1 and the Pass Equivalence Graf.

\texttt{PASS(n+1, Completer Loop @ Marpa)}. By the assumption for
Sublemma 1 and the Pass Equivalence Graf.

\texttt{(S{[}i+f{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom of Completer
Loop(p)}. By the previous 2 grafs and the Earley Item Induction Property

\noun{This proves the induction}.

\noun{This proves the sublemma}.

\noun{Note}: The sublemma does most of the work of this Lemma. In
statements that follow, I show \texttt{(S{[}i{]} @ AH2002 == S{[}i{]}
@ Marpa)} holds as we proceed through the Earley Set Processing. To
do this, I use the sublemma for the portion of the Earley Set Processing
where \texttt{S{[}i{]}} might change. Where \texttt{S{[}i{]}} remains
unchanged in both algorithms, this is established by referring to
the pseudocode.

(\texttt{S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Start of Earley
Set i Processing}. From the assumption in the Statement of the Lemma,
substituting \texttt{S{[}i{]}} for \texttt{S}.

\texttt{((S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ Top of Completer
Loop(0)) @ Earley Set i Processing}. From the pseudocode and the previous
graf. {[} Top Identical Graf. {]}

Let \texttt{z} be the pass such that \texttt{(AH2002 @ End of Completer
Loop == AH2002 @ Bottom of Completer Loop(z)) @ Earley Set i Processing}.
By the AH2002 Halting Axiom there is such a pass.

\texttt{(S{[}i{]} @ AH2002 @ Bottom of Completer Loop(z) == S{[}i{]}
@ Marpa @ Bottom of Completer Loop(z)) @ Earley Set i Processing}.
By the previous two statements and sublemma. Pass \texttt{z} exists
for AH2002 by definition. That it exists for Marpa will be shown by
the sublemma.

\texttt{(S{[}i{]} @ Marpa @ End of Completer Loop == S{[}i{]} @ Marpa
@ Bottom of Completer Loop(z)) @ Earley Set i Processing}. By the
last two statements and the loop definitions. If at pass z the Earley
sets are identical, and the Earley set in one algorithm ends the loop,
it must also end the loop in the other algorithm at pass z.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ End of Completer
Loop @ Earley Set i Processing}. By the three statements just previous.

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ End of Earley Set
i Processing}. From the pseudocode.

\noun{This proves the Lemma.}


\subsection{The Scanner Lemma}

\noun{Statement of the Lemma}:

Assume \texttt{PASS(p,} \texttt{Scanner Loop}).

Assume that \texttt{((state, parent) @ AH2002 == (state, parent) @
Marpa) @ Top of Scanner Loop}(p).

Assume that \texttt{(i @ AH2002 == i @ Marpa) @ Top of Scanner Loop(p)}.

Assume \texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of
Scanner Loop(p)}.

Then \texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Bottom
of Scanner Loop(p).}

\noun{Starting the proof}:

\texttt{0 <= p <= SIZE(S{[}i{]} @ End of Scanner Loop)}. From as assumption
for the lemma and the definition of \texttt{PASS}.

The proof proceeds in two cases -- the Final Case (\texttt{p == SIZE(S{[}i{]}
@ End of Scanner Loop)}) and the Kernel Case (\texttt{0 <= p < SIZE(S{[}i{]}
@ End of Scanner Loop)}).

\noun{Kernel case}:

\texttt{((state, parent, i) @ AH2002 == (state, parent, i) @ Marpa)
@ Top of Scanner Loop(p)}. By assumption in the statement of the lemma.

\texttt{((state, parent, i) @ AH2002 == (state, parent, i) @ Marpa)
@ Top of Scanner @ Scanner Loop(p)}. From the graf just previous and
the psuedocode. These variables are never changed between \texttt{Top
of Scanner Loop(p)} and \texttt{Top of Scanner @ Scanner Loop(p)}.
{[} \noun{Input Variables Graf {]}}

\texttt{G} $\in$ \texttt{(x, GOTO)} $\Rightarrow$ \texttt{(G @ Marpa
== G @ AH2002) @ Top of Scanner @ Scanner Loop(p)}. By the Globals
Lemmas. {[} \noun{Globals Graf {]}}

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of Scanner
Loop(p)}. By an assumption in the statement of the lemma.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of Scanner
@ Scanner Loop(p)}. From the graf just previous and inspection of
the pseudocode. This Earley set is never changed outside of Scanner.
{[} \noun{Next Earley Set Graf {]}}

\texttt{S{[}i+1{]} @ AH2002 @ Bottom of Scanner @ Scanner Loop(p)
== S{[}i+1{]} @ Marpa @ Bottom of Scanner @ Scanner Loop(p)}. From
the pseudocode, the Input Variables Statement, the Globals Statement
and Destination Statement. Inspection of Scanner shows that \texttt{S{[}i+1{]}
@ End of Scanner} is a deterministic function. Inputs of the function
are \texttt{S{[}i+1{]}} itself, the \texttt{Scanner}'s input variables
and the globals \texttt{x} and \texttt{GOTO}. These inputs are the
same in both algorithms by the Next Earley Set Graf, the Input Variables
Graf and the Globals Graf. The \texttt{Scanner} is also the same in
both algorithms. Therefore the result must be the same.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom of Scanner
Loop(p)}. From the pseudocode. This Earley set is never changed between
\texttt{Bottom of Scanner @ Scanner Loop(p)} and \texttt{Bottom of
Scanner Loop(p)}.

\noun{This proves the Kernel case}.

\noun{Final case}:

\noun{Assumption for this case}: \texttt{FINAL(p, Scanner Loop).}

\texttt{(S{[}i+1{]} @ Marpa == S{[}i+1{]} @ AH2002) @ Bottom of Scanner
Loop(p)}. From an assumption in the statement of this lemma, the graf
just previous and the loop definitions. The final pass of a loop is
always a no-op.

\noun{This proves the final case}.

With the two cases proved, the Lemma is shown.

QED.


\subsection{The Scanner Loop Lemma}

\noun{Statement of the Lemma}:

Assume that \texttt{PASS(p,} \texttt{Scanner Loop @ Earley Set i Processing}).

Assume that \texttt{(S @ AH2002 == S @ Marpa) @ End of Earley Set
i Processing}.

Assume \texttt{(i @ Marpa == i @ AH2002) @ Earley Set Processing}.

Assume \texttt{(S{[}i+1{]} @ Marpa == S{[}i+1{]} @ AH2002) @ Top of
Scanner Loop(p)}.

Then \texttt{(S{[}i+1{]} @ Marpa == S{[}i+1{]} @ AH2002) @ Bottom
of Scanner Loop (p)}.

\noun{Proof:}

\texttt{(S{[}i{]} @ AH2002 == S{[}i{]} @ Marpa) @ End of Earley Set
i Processing}. By an assumption for the lemma, with S{[}i{]} substituted
for S.

\texttt{(S{[}i{]}{[}p{]} @ AH2002 == S{[}i{]}{[}p{]} @ Marpa) @ End
of Earley Set i Processing}. From the statement just previous. p exists
in both algorithms because S{[}i{]} is the domain of Scanner Loop,
and by assumption for the lemma, p is a pass.

\texttt{End of Scanner Loop @ Earley Set i Processing == End of Earley
Set i Processing}. From the pseudocode. Recall that \texttt{Combined
Loop @ AH2002 == Scanner Loop @ AH2002}.

\texttt{(S{[}i{]}{[}p{]} @ AH2002 == S{[}i{]}{[}p{]} @ Marpa) @ End
of Scanner Loop @ Earley Set i Processing}. From the previous 2 grafs.

\texttt{(S{[}i{]}{[}p{]} @ Top of Scanner Loop(p) == S{[}i{]}{[}p{]}
@ End of Scanner Loop) @ Earley Set i Processing}. S{[}i{]}{[}p{]}
exists at \texttt{Top of Scanner Loop(p)} by the definition of PASS.
It is identical to S{[}i{]}{[}p{]} at End of Scanner Loop because
worklist items, once they exist, are never changed or deleted.

\texttt{(S{[}i{]}{[}p{]} @ AH2002 == S{[}i{]}{[}p{]} @ Marpa) @ Top
of Scanner Loop(p) @ Earley Set i Processing}. From the previous 2
grafs.

\texttt{((state, parent) == S{[}i{]}{[}p{]}) @ Top of Scanner Loop(p)
@ Earley Set i Processing}. Because (state, parent) and S{[}i{]}{[}p{]}
are two different names for the loop variable, they must be identical.

\texttt{((state, parent) @ AH2002 == (state, parent) @ Marpa) @ Top
of Scanner Loop(p) @ Earley Set i Processing}. From the previous two
statements.

\texttt{(i @ AH2002 == i @ Marpa) @ Top of Scanner Loop (p)}. From
an assumption in the statement of this lemma.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of Scanner
Loop(p)}. An assumption in the statement of this lemma.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Bottom of Scanner
Loop(p)}. By the Scanner Lemma with p set to its value in this Lemma,
and by the three statements just previous.

\noun{This proves the lemma.}


\subsection{The Next Earley Set Lemma}

\noun{Statement of the Lemma}:

Assume \texttt{SIZE(S{[}i{]} @ AH2002) == SIZE(S{[}i{]} @ Marpa) @
End of Earley Set i Processing}.

Assume \texttt{0 <= i <= SIZE(x)}.

Assume \texttt{(S @ AH2002 == S @ Marpa) @ Start of Earley Set i Processing}.

Then, \texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ End of
Earley Set i Processing}.

\noun{Starting the Proof}:

\noun{Sublemma}:

Assume \texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of
Scanner Loop(0)}.

Assume \texttt{p <= SIZE(S{[}i+f{]}) @ AH2002 @ Bottom of Scanner
Loop(p)} or \texttt{p <= SIZE(S{[}i+f{]}) @ Marpa @ Bottom of Scanner
Loop(p)}.

Then \texttt{(S{[}i+f{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom
of Scanner Loop(p)}.

\noun{Note}: In this sublemma, it is assumed in the statement of the
sublemma that pass p exists for one algorithm, but not for the other.
Loop operations are not defined for non-existent passes, and it must
be proved that pass p exists for both AH2002 and Marpa.

\noun{Starting the proof}:

\noun{Induction Property}: \texttt{(S{[}i+f{]} @ A == S{[}i+f{]} @
Marpa) @ Bottom of Scanner Loop(p)}, where the induction is on \texttt{p}.

\noun{Induction} \noun{Basis}: \texttt{(S{[}i+f{]} @ A == S{[}i+f{]}
@ Marpa) @ Bottom of Scanner Loop(0)}. By the assumptions in the statement
of Sublemma 1 and the Earley Item Induction Property Lemma with p
== 0. Pass 0 always exists for every loop, by definition.

\noun{This proves the induction basis.}

\noun{Induction hypothesis}:

Assume \texttt{(S{[}i+f{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom
of Scanner Loop(n)}.

\noun{Inductive step}: 

Note: Next it must be shown that \texttt{n+1} is a pass in both \texttt{AH2002}
and \texttt{Marpa}. 

\noun{Subsublemma}: Let \texttt{(A,B)} be either \texttt{(AH2002,
Marpa) }or \texttt{(Marpa, AH2002)}.

Assume \texttt{PASS(n+1, Scanner Loop @ A)}.

Then \texttt{PASS(n+1, Scanner Loop @ B)}.

\texttt{SIZE(S{[}i{]} @ AH2002) == SIZE(S{[}i{]} @ Marpa) @ End of
Earley Set i Processing}. From an assumption for this lemma.

\texttt{(S{[}i{]} @ A == S{[}i{]} @ B) @ End of Earley Set i Processing.}
By the previous, substituting A and B, and transposing the equality
if necessary.

\texttt{(S{[}i{]} @ A == S{[}i{]} @ B) @ End of Scanner Loop.} From
the previous graf and the pseudocode. Nothing between \texttt{End
of Scanner Loop} and \texttt{End of Earley Set i Processing} changes
\texttt{S{[}i{]}. {[} }\noun{End of Scanner Loop Equality Graf} {]}

\texttt{0 <= n < SIZE(DOMAIN(Scanner Loop) @ Bottom of Scanner Loop(n)
@ A))}. By assumption for the sublemma and the definition of \texttt{PASS}.

\texttt{0 <= n+1 <= SIZE(S{[}i{]} @ Bottom of Scanner Loop(n) @ A)}.
By the previous graf, the loop definitions. By the definition of \texttt{DOMAIN},
\texttt{DOMAIN(Scanner Loop) == S{[}i{]}}.

\texttt{SIZE(S{[}i{]} @ Bottom of Scanner Loop(n) @ A) <= SIZE(S{[}i{]}
@ End of Scanner Loop @ A)}. By the loop definitions. End of Scanner
Loop is at or after Bottom of Scanner Loop(n) and the loop domain
never decreases in size.

\texttt{0 <= n+1 <= SIZE(S{[}i{]} @ End of Scanner Loop @ A)}. From
the previous 2 grafs.

\texttt{0 <= n+1 <= SIZE(S{[}i{]} @ End of Scanner Loop @ B)}. From
the previous and the \texttt{End of Scanner Loop Equality Graf}.

\texttt{PASS(n+1, Scanner Loop @ B)}. From the previous graf and definition
of \texttt{PASS}.

\noun{This proves the subsublemma}.

\texttt{PASS(n+1, Scanner Loop @ AH2002) $\Rightarrow$ PASS(n+1,
Scanner Loop @ Marpa)}. By the subsublemma, with \texttt{(A, B) :=
(AH2002, Marpa)}.

\texttt{PASS(n+1, Scanner Loop @ Marpa) $\Rightarrow$ PASS(n+1, Scanner
Loop @ AH2002)}. By the subsublemma, with \texttt{(A, B) := (Marpa,
AH2002)}.

\texttt{PASS(n+1, Scanner Loop @ AH2002) $\Leftrightarrow$PASS(n+1,
Scanner Loop @ Marpa)}. By the previous two grafs. {[} \noun{Pass
Equivalence Graf} {]}

\texttt{PASS(n+1, Scanner Loop @ AH2002)}. By the assumption for the
sublemma and the Pass Equivalence Graf.

\texttt{PASS(n+1, Scanner Loop @ Marpa)}. By the assumption for the
sublemma and the Pass Equivalence Graf.

\texttt{(S{[}i+f{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Top of Scanner
Loop(p+1)}. By the loop definitions and the induction hypothesis.
The two grafs just previous show that pass \texttt{n+1} exists for
both algorithms.

\texttt{(S{[}i+f{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom of Scanner
Loop(p+1)}. By the previous graf and the Scanner Loop Lemma.

\noun{This proves the induction}.

\noun{This proves sublemma}.

\noun{Note}: The sublemma does most of the work of this lemma. In
statements that follow, I show \texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]}
@ Marpa)} holds as we proceed through the \texttt{Earley Set Processing}.
To do this, I use the sublemma for the portion of the \texttt{Earley
Set Processing} where \texttt{S{[}i+1{]}} might change. In most cases
\texttt{S{[}i+1{]} }remains unchanged in both algorithms, as is clear
from the pseudocode.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Start of Earley
Set i Processing}. From the assumption in the Statement of the Lemma,
substituting \texttt{S{[}i+1{]}} for \texttt{S}.

\texttt{((S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ Top of Scanner
Loop(0)) @ Earley Set i Processing}. From the pseudocode and the previous
graf. {[} \noun{Top Identical Graf} {]}

Let \texttt{z} be the pass such that \texttt{(AH2002 @ End of Scanner
Loop == AH2002 @ Bottom of Scanner Loop(z)) @ Earley Set i Processing}.
By the AH2002 Halting Axiom there is such a pass.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+f{]} @ Marpa) @ Bottom of Scanner
Loop(z) @ Earley Set i Processing}. By the previous two statements
and sublemma. Pass \texttt{z} exists for \texttt{AH2002} by definition.
That it exists for \texttt{Marpa} will be shown by the sublemma.

\texttt{(S{[}i+1{]} @ Marpa @ End of Scanner Loop == S{[}i+1{]} @
Marpa @ Bottom of Scanner Loop(z)) @ Earley Set i Processing}. By
the last two statements and the loop definitions. If at pass \texttt{z}
the Earley sets are identical, and the Earley set in one algorithm
ends the loop, it must also end the loop in the other algorithm at
pass \texttt{z}.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ End of Scanner
Loop @ Earley Set i Processing}. By the three statements just previous.

\texttt{(S{[}i+1{]} @ AH2002 == S{[}i+1{]} @ Marpa) @ End of Earley
Set i Processing}. From the pseudocode.

\noun{This proves the Lemma.}


\subsection{Earley Set Processing Lemma}

\noun{Statement of the Lemma}:

Assume for \texttt{(S @ Marpa == S @ AH2002) @ Start of Earley Set
n Processing}.

Assume also, that \texttt{0 < n < SIZE(x)}.

Then \texttt{(S @ Marpa == S @ AH2002) @ End of Earley Set n Processing}.

\noun{Starting the Proof}:

For all \texttt{x}, \texttt{(x == n or x == n+1 or S{[}x{]} @ Marpa
== S{[}x{]} @ AH2002) @ End of Earley Set n Processing}. From the
Globals Lemma, and the assumption in the Statement of the Lemma.

\texttt{(S{[}n{]} @ Marpa == S{[}n{]} @ AH2002) @ End of Earley Set
n Processing}. From the Current Earley Set Lemma \texttt{(i := n)},
and the assumptions in the Statement of the Lemma.

\texttt{(S{[}n+1{]} @ Marpa == S{[}n+1{]} @ AH2002) @ End of Earley
Set n Processing}. From the Next Earley Set Lemma \texttt{(i := n)},
the previous graf, and the assumptions in the Statement of the Lemma.

For all x, \texttt{(S{[}x{]} @ Marpa == S{[}x{]} @ AH2002) @ End of
Earley Set n Processing}. Combining the three statements just previous.

\texttt{(S @ Marpa == S @ AH2002) @ End of Earley Set n Processing}.
From the previous statement.

QED.


\subsection{AH2002 and Marpa are Equivalent}

We regard AH2002 and Marpa as equivalent if, given the Globals Assumption,
the Earley sets are identical at the end of each algorithm. The Globals
Assumption has been assumed throughout this proof.

\noun{Statement of the Theorem}: \texttt{S @ End of AH2002 == S @
End of Marpa}.

\noun{lemma}: Assume \texttt{0 < s < SIZE(x)}.

Then \texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Step s}.

\noun{Starting the Proof}: The proof is by induction on the Recognizer
Steps.

\noun{Induction Property}: \texttt{(S @ AH2002 == S @ Marpa) @ End
of Recognizer Step t}, where the induction is on \texttt{t}.

\noun{Induction Base}:

\texttt{(S @ AH2002 == S @ Marpa) @ Start of Recognizer Step 0}. By
the Globals Lemmas.

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Step 0}. By
the just previous statement, the Globals Lemmas and the Earley Sets
Processing Lemma with \texttt{n := 0}. There will always be an Earley
set 0 and therefore by the definition of the Recognizer, a Recognizer
Step 0.

\noun{Induction Hypothesis}:

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Step t}.

\noun{Inductive Step}:

\texttt{0 <= t+1 < SIZE(x)}. From the statement of the Sublemma. {[}
\noun{Induction Limit Graf {]}}

\texttt{(S @ AH2002 == S @ Marpa) @ Start of Recognizer Step t+}1.
By the definition of the Recognizer, the Induction Hypothesis and
the Induction Limit Graf.

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Step t+1}. By
the just previous statement, the Induction Limit Graf, and the Earley
Set Processing Lemma with \texttt{(n := s+1)}.

\noun{This completes the induction}.

\noun{This proves the lemma}.

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Loop(SIZE(x)
- 1)}. By the lemma with \texttt{s := SIZE(x)-1}.

\texttt{(S @ AH2002 == S @ Marpa) @ Start of Recognizer Loop(SIZE(x))}.
By the graf just previous and the loop definitions. SIZE(x) is a pass
by the definition of the Recognizer.

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer Loop(SIZE(x))}.
By the loop definitions and the previous graf. SIZE(x) is the final
pass of Recognizer Loop, by its definition. The final pass is a no-op,
so the values of S will not change in either algorithm.

\texttt{(S @ AH2002 == S @ Marpa) @ End of Recognizer}. By the definition
of the Recognizer and the previous graf.

\texttt{S @ End of AH2002 == S @ End of Marpa}. By the graf just previous
and by the definition of the two algorithms.

\noun{QED}.
\end{document}
